---
title: "Darwin Core mapping"
subtitle: "For: Alien species occurrences across Europe: Data collected during a pan-European Bioblitz"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://docs.google.com/spreadsheets/d/1LoysjuWllDQRtpaVIPIhGN95bV_E8jhSnqe0tOsDx3o/edit?pli=1#gid=1668158497).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
library(rgbif)
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1LoysjuWllDQRtpaVIPIhGN95bV_E8jhSnqe0tOsDx3o/edit?pli=1#gid=1668158497).

Read the relevant worksheet (published as csv):

```{r read_source_data}
input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vT5kQzv0ngSRDp_m-4WzL9PRddvtFDClbCes_oix54MtFCrb4rFzUPp-c8NqTHtnOY0mBpUOStUh7hr/pub?gid=1668158497&single=true&output=csv")
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_data, here("data", "raw", "bioblitz_data_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```
# Preprocessing

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

One records `id = 13057` is lacking coordinates. For some of the Czech records, no identifier is available. Removing the record for now:

```{r}
input_data <- input_data %>% filter(!is.na(id)) %>% filter(id != 13057)
```

Check scientific names:

```{r}
parsed_names <- input_data %>%
  distinct(sps) %>%
  pull() %>% # Create vector from dataframe
  parsenames() # An rgbif function
```

Show scientific names with nomenclatural issues, i.e. not of `type = SCIENTIFIC` or that could not be fully parsed. Note: these are not necessarily incorrect.

```{r}
parsed_names %>%
  select(scientificname, type, parsed, parsedpartially, rankmarker) %>%
  filter(!(type == "SCIENTIFIC" & parsed == "TRUE" & parsedpartially == "FALSE"))
```

# Darwin Core mapping

## Occurrence Core

Map the data to [Darwin Core Occurrence](https://rs.gbif.org/core/dwc_occurrence_2022-02-02.xml).

Create `occurrence`:

```{r}
occurrence <- input_data
```

### language

```{r}
occurrence %<>% mutate(dwc_language = "en")
```

### license

```{r}
occurrence %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
occurrence %<>% mutate(dwc_rightsHolder = "") 
```

### accessRights

```{r}
occurrence %<>% mutate(dwc_accessRights = "https://www.inbo.be/en/norms-data-use") 
```

### datasetID

```{r}
occurrence %<>% mutate(dwc_datasetID = "") 
```

### institutionCode

```{r}
occurrence %<>% mutate(dwc_institutionCode = "") 
```

### datasetName

```{r}
occurrence %<>% mutate(dwc_datasetName = "Alien species occurrences across Europe: Data collected during a pan-European Bioblitz") 
```

## eventDate

```{r}
occurrence %<>% mutate(dwc_eventDate = obs_date) 
```

## country

```{r}
occurrence <-
  occurrence %>% mutate(dwc_country = case_when(
    region == "CanaryIslands" | region == "Catalonia" ~ "Spain",
    TRUE ~ region))
```

## countryCode

```{r}
occurrence <- 
  occurrence %>% mutate(dwc_countryCode = recode(dwc_country,
    "Austria" = "ISO-3166:",
    "Belgium" = "ISO-3166:",
    "Bulgaria" = "ISO-3166:",
    "Croatia" = "ISO-3166:",
    "Cyprus" = "ISO-3166:",
    "Czechia" = "ISO-3166:",
    "Greece" = "ISO-3166:",
    "Italy" = "ISO-3166:",
    "Poland" = "ISO-3166:",
    "Portugal" = "ISO-316:",
    "Spain" = "ISO-316:",
    "Turkey" = "ISO-316:"
))
                                             
```

## locality

```{r}
occurrence <-
  occurrence %>% mutate(dwc_locality = case_when(
    region == "Bulgaria" ~ "Pancharevo Lake",
    region_long == "CanaryIslands" ~ "Canary Islands",
    region_long == "El Hierro" ~ "Canary Islands - El Hierro",
    region_long == "Fuerteventura" ~ "Canary Islands - Fuerteventura",
    region_long == "Gran Canaria" ~ "Canary Islands - Gran Canaria",
    region_long == "La Gomera" ~ "Canary Islands - La Gomera",
    region_long == "La Graciosa" ~ "Canary Islands - La Graciosa",
    region_long == "La Palma" ~ "Canary Islands - La Palma",
    region_long == "Lanzarote" ~ "Canary Islands - Lanzarote",
    region_long == "Tenerife" ~ "Canary Islands - Tenerife",
    region_long == "Catalonia" ~ "Catalonia",
    region == "Greece" ~ "Kaisariani Forest",
    region == "Italy" ~ "Sardinia",    
    region == "Portugal" ~ "Ria de Aveiro",
    region == "Turkey" ~ "Ankara & Antalya",
    TRUE ~ ""))
```

## decimalLatitude, decimalLongitude, geodeticDatum

```{r}
occurrence <- 
  occurrence %>% 
    mutate(dwc_decimalLatitude = lat) %>% 
    mutate(dwc_decimalLongitude = lon) %>% 
    mutate(dwc_geodeticDatum = "WGS84")
```

## occurrenceID

```{r}
occurrence %<>% mutate(dwc_occurrenceID = id) 
```

## recordedBy

```{r}
occurrence %<>% mutate(dwc_recordedBy = author) 
```

## establishmentMeans

```{r}
occurrence <- 
  occurrence %>% mutate(dwc_establishmentMeans = case_when(
    status_native_alien_ias == "alien"  |
    status_native_alien_ias == "alien-naturalized" |
    status_native_alien_ias == "IAS" |
    status_native_alien_ias == "invasive" ~ "introduced",
    status_native_alien_ias == "native" ~ "native",
    TRUE ~ ""
    ))
```

## degreeOfEstablishment

```{r}
occurrence <- 
  occurrence %>% mutate(dwc_degreeOfEstablishment = case_when(
    status_native_alien_ias == "alien-naturalized" ~ "established",
    status_native_alien_ias == "cultivated (see comment)" ~ "cultivated",
    status_native_alien_ias == "IAS" | status_native_alien_ias == "invasive" ~ "invasive",
    status_native_alien_ias == "native" ~ "native",
    status_native_alien_ias ==  "planted and also escaping" ~ "released",
    TRUE ~ ""
  ))
```

## scientific name

```{r}
occurrence %<>% mutate(dwc_scientificName = sps)
```

## vernacularName

```{r}
occurrence %<>% mutate(dwc_vernacularName = sps_comm)
```

# Post-processing

Only keep the Darwin Core columns:

```{r}
occurrence %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(occurrence) <- str_remove(colnames(occurrence), "dwc_")
```

Preview data:

```{r}
occurrence %>% head()
```

Save to CSV:

```{r}
write_csv(occurrence, here("data", "processed", "occurrence.csv"), na = "")
```






